---
title: "Regression Models Course Project"
author: "William Matthews"
date: "March 10, 2018"
output: 
  pdf_document: 
    fig_caption: yes
    fig_height: 5.5
    fig_width: 6.5
keep_md: yes
---
## Executive Summary
We will be reviewing the mtcars dataset and exploring the relationship between its set of variables and
miles per gallon (MPG) as the outcome. Of particular interest are the following questions:

1. Is an automatic or manual transmission better for mpg?
2. Can I quantify the mpg difference between automatic and manual transmission?

## Conclusions:

Based on our analysis cars with manual transmission get higher mpg than those with automatic transmission. 
The difference is about 7.2 mpg. However, transmission type only accounts for about 36% of the variability
in mpg. As a result we looked at the other variables in the mtcars dataset to see what affect they had on
mpg. 
 
The other variables from mtcars that had the highest affect on mpg were hp, cyl, and wt. 
 
When adjusting for hp, cyl, & wt manual transmission increased mpg by about 1.8 over automatic transmission.
Mpg decreased by about 2.5 for every 1000 lb. increase in wt (adjusted for hp, cyl, & am).
When cyl (the number of cylinders) increases from 4 to 6 to 8, mpg will decrease by 3 and 2.2 respectively
(adjusted for hp, wt, & am). 

```{r include=FALSE}
# Load tidyverse package
library(tidyverse)
```

```{r}

# Load mtcars dataset
data(mtcars) 
```

## Exploratory Data Analysis
```{r}
# Use dim() to obtain the dimensions of the data frame
dim(mtcars) 
```

```{r}
# Use head() to obtain the first n observations of the dataset.
head(mtcars, 3)
```

```{r}
# Use the str() function to return the structure of the mtcars dataset. 
str(mtcars)
```

```{r}
# Display the correlation between mpg and the other variables in mtcars
cor(mtcars)[1,]
```

The variables with the highest correlation to mpg are cyl, disp, hp, & wt

```{r}
# Recode selected numeric variables as factors
mtcars$cyl <- factor(mtcars$cyl)
mtcars$am <- factor(mtcars$am, labels = c("Automatic", "Manual")) # 0 = automatic, 1 = manual
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
```

## Regression Analysis
```{r}
# Aggregate the mpg by transmission type (auto, manual)
aggregate(mpg~am, data = mtcars, mean)
```

The data indicates that on average mpg is higher with manual transmissions. 

## Statistical Inference

```{r}
# Quantify the difference in mpg for the am variable with a t-test.
t.test(mpg ~ am, data = mtcars)
```

The p-value = 0.001374, which is less than .05, indicating that this is a significant difference and thus reject the null hypothesis that automatic and manual transmissions have the same effect on mpg.  

### Model selection
#### Linear models

```{r}
# Use lm function to fit a linear model with mpg as the outcome and am as the predictor.
fit <- lm(mpg~am, data = mtcars)

# View summary of lm fit.
summary(fit)
```

The lm summary shows that mpg increases by 7.245 for manual transmission. 
The R-squared:  0.3598 indicates that this model explains only 34% of the variance of mpg.  

#### Multivariate linear model
```{r}
# Use the lm function to fit the linear model with mpg as the outcome and the other variables as 
# predictors.
fit1 <- lm(mpg ~ ., data = mtcars)
```

Use the step fuction to select a formula-based model by AIC of the variables that have the highest correlation to mpg, ie the "best" model.
mv_fit <- step(fit1, direction = "both")
```{r message=FALSE, include=FALSE}
mv_fit <- step(fit1, direction = "both")

```

The model with the lowest AIC and thus fit is mpg ~ cyl + hp + wt + am

```{r}
# View summary of mv_fit model
summary(mv_fit)
```

When accounting for the other variables (cyl, hp, wt) manual transmission increases mpg by 1.8. 
The R^2 value indicates tht 86.59% of the variance is explained by the model. 

```{r}
# Use anova function to compare the base model against the best model.
anova(fit, mv_fit)
```

```{r}
# Test the confidence of the model.
confint(mv_fit)
```

We can say with 95% confidence that the variables correlations are within the ranges listed. 

### Residual

The Residuals vs Fitted plot (Appendix - Plot 3) shows that the points are randomly distributed indicating independence.
The Normal Q-Q plot shows that the distribution is generally normal because the points mostly fall on the normal line.
The Scale-Location plot shows the points scattered in a constant pattern indicating a constance variance condition.
The Residuals vs Leverage plot shows some outliers

\newpage

## Appendix

### Plot 1: Plot mpg by transmission type with mpg on the y axis and transmission type on the x axis.
```{r}
plot(mpg~am, data = mtcars)
```

\newpage

### Plot 2: Use the pairs function to plot a matrix of the relationship betwen mpg and the other variables.
```{r}
pairs(mpg~., data = mtcars)
```

\newpage

### Plot 3: View the residual plots for multivariate regression model and compute regression diagnostic of model to uncover outliers.
```{r}
par(mfrow = c(2,2))
plot(mv_fit)
```